Concurrent Programming in Harmony

### 2021-12-17 book/paper.tex
Some small updates

Method \texttt{processor}(\textit{self}, \textit{succ}) is the
code for a processor with identifier \textit{self} and successor
\textit{succ}.
It starts simply by sending its own identifier to its successor.
The processor then loops until it discovers the identifier of the
leader in the second phase of the protocol.
A processor waits for a message using the Harmony
\textbf{atomically when exists} statement.
This statement takes the form \textbf{atomically when exists} $v$ \textbf{in} $s$: \emph{statement block}
where $s$ is a set and $v$ is variable that is bound to an element of $s$.
The properties of the statement are as follows:
\begin{itemize}
\item it waits until $s$ is non-empty;
\item it is executed atomically;
\item $v$ is selected non-deterministically, like in the \texttt{choose} operator.
\end{itemize}
\index{select statement}%

@@

Note the use of the \textbf{possibly} statement in this code.
It is similar to \textbf{assert}, but where \texttt{assert} checks that
the predicate holds in \emph{all} executions, \texttt{possibly} checks
that the predicate holds in at least one execution.
The purpose is to check that the code doesn't simply always abort---there
exist executions where the code commits.
In practice, hopefully \emph{most} executions commit, but Harmony is not
capable of probabilistic reasoning.

@@

The \texttt{possibly} statement checks if there exist executions where 0 is
decided and if there exist executions where 1 is decided.  Doing so is useful
because the assertions would certainly not fail if no decision is ever made.
Also, it is nice to know that both 0 and 1 can be decided.

@@

\vspace{1em}
\begin{tabular}{|l|l|}
\hline
$e$ & $e$ is an expression\\
\hline
\textit{lv} = [lv =]... $e$ & \textit{lv} is an lvalue and $e$ is an expression\\
\hline
\textit{lv} \texttt{[op]=} $e$ & \texttt{op} is one of \texttt{+}, \texttt{-},
\texttt{*}, \texttt{/}, \texttt{//}, \texttt{\%},
\texttt{\&}, \texttt{|}, \texttt{\^},
\texttt{and}, \texttt{or}\\
\hline
\texttt{assert $b$ [, $e$]} & $b$ is a boolean.  Optionally report value of expression $e$\\
\hline
\texttt{await $b$} & $b$ is a boolean\\
\hline
\texttt{const a = $e$} & \texttt{a} is a bound variable, $e$ is a constant expression\\
\hline
\texttt{def m $a$: S} & \texttt{m} is an identifier, $a$ a bound variable\\
\hline
\texttt{del} \textit{lv} & delete\\
\hline
\texttt{for $a[:b]$ in $e$ [where $c$]: S} & $a$ and $b$ are bound variables, $e$ is a set or dictionary \\
\hline
\texttt{from m import ...} & \texttt{m} identifies a module \\
\hline
\texttt{go $c$ $e$} & $c$ is a context, $e$ is an expression \\
\hline
\texttt{if $b$: S else: S} & $b$ is a boolean, \texttt{S} is a list of statements\\
\hline
\texttt{import m, ...} & \texttt{m} identifies a module \\
\hline
\texttt{invariant e} & \texttt{e} is an invariant\\
\hline
\texttt{let $a$ = $e$: S} & $a$ is a bound variable, $e$ is an expression\\
\hline
\texttt{pass} & do nothing\\
\hline
\texttt{possibly $b$ [, $b$, ...]} & each $b$ is a boolean \\
\hline
\texttt{select $a$ in $e$: S} & $a$ is a bound variable, $e$ is a set,
                            \texttt{S} a list of statements\\
\hline
\texttt{sequential v, ...} & \texttt{v} has sequential consistency \\
\hline
\texttt{spawn [eternal] m $e$ [, $t$]} & \texttt{m} is a method,
$e$ is an expression, $t$ is the thread-local state \\
\hline
\texttt{trap m $e$} & \texttt{m} is a method and $e$ is an expression \\
\hline
\texttt{var $v$ = $e$} & $v$ is a new variable, $e$ is an expression\\
\hline
\texttt{when $b$: S} & $b$ is a boolean, \texttt{S} a list of statements\\
\hline
\texttt{when exists $a$ in $e$: S} & $a$ is a bound variable, $e$ is an expression\\
\hline
\texttt{while $b$: S} & $b$ is a boolean, \texttt{S} a list of statements\\
\hline
\end{tabular}

@@

\subsection*{\textbf{possibly}}

The statement \textbf{possibly $b_1, b_2, ...$} atomically evaluates all
predicates $b_i$.
At completion, Harmony reports which of the predicates never held.
While \texttt{assert} statements check that nothing ever goes wrong,
\texttt{possibly} statements can check that certain things sometimes
go right.

>>>>

Method \texttt{processor}(\textit{self}, \textit{succ}) is the
code for a processor with identifier \textit{self} and successor
\textit{succ}.
It starts simply by sending its own identifier to its successor.
The processor then loops until it discovers the identifier of the
leader in the second phase of the protocol.
A processor waits for a message using the Harmony
\textbf{atomically when exists} statement.
This statement takes the form \textbf{atomically when exists} $v$ \textbf{in} $s$: \emph{statement block}
where $s$ is a set and $v$ is variable that is bound to an element of $s$.
The properties of the statement are as follows:
\begin{itemize}
\item it waits until $s$ is non-empty;
\item it is executed atomically;
\item $v$ is selected non-deterministically, like in the \texttt{choose} operator.
\end{itemize}

@@

\begin{comment}
Note the use of the \textbf{possibly} statement in this code.
It is similar to \textbf{assert}, but where \texttt{assert} checks that
the predicate holds in \emph{all} executions, \texttt{possibly} checks
that the predicate holds in at least one execution.
The purpose is to check that the code doesn't simply always abort---there
exist executions where the code commits.
In practice, hopefully \emph{most} executions commit, but Harmony is not
capable of probabilistic reasoning.
\end{comment}

@@

\begin{comment}
The \texttt{possibly} statement checks if there exist executions where 0 is
decided and if there exist executions where 1 is decided.  Doing so is useful
because the assertions would certainly not fail if no decision is ever made.
Also, it is nice to know that both 0 and 1 can be decided.
\end{comment}

@@

\vspace{1em}
\begin{tabular}{|l|l|}
\hline
$e$ & $e$ is an expression\\
\hline
\textit{lv} = [lv =]... $e$ & \textit{lv} is an lvalue and $e$ is an expression\\
\hline
\textit{lv} \texttt{[op]=} $e$ & \texttt{op} is one of \texttt{+}, \texttt{-},
\texttt{*}, \texttt{/}, \texttt{//}, \texttt{\%},
\texttt{\&}, \texttt{|}, \texttt{\^},
\texttt{and}, \texttt{or}\\
\hline
\texttt{@l: S} & \texttt{l} is a label \\
\hline
\texttt{assert $b$ [, $e$]} & $b$ is a boolean.  Optionally report value of expression $e$\\
\hline
\texttt{await $b$} & $b$ is a boolean\\
\hline
\texttt{const a = $e$} & \texttt{a} is a bound variable, $e$ is a constant expression\\
\hline
\texttt{def m $a$: S} & \texttt{m} is an identifier, $a$ a bound variable\\
\hline
\texttt{del} \textit{lv} & delete\\
\hline
\texttt{for $a[:b]$ in $e$ [where $c$]: S} & $a$ and $b$ are bound variables, $e$ is a set or dictionary \\
\hline
\texttt{from m import ...} & \texttt{m} identifies a module \\
\hline
\texttt{go $c$ $e$} & $c$ is a context, $e$ is an expression \\
\hline
\texttt{if $b$: S else: S} & $b$ is a boolean, \texttt{S} is a list of statements\\
\hline
\texttt{import m, ...} & \texttt{m} identifies a module \\
\hline
\texttt{invariant e} & \texttt{e} is an invariant\\
\hline
\texttt{let $a$ = $e$: S} & $a$ is a bound variable, $e$ is an expression\\
\hline
\texttt{pass} & do nothing\\
\hline
\texttt{print $b$ [, $b$, ...]} & each $b$ is a boolean \\
\hline
\texttt{sequential v, ...} & \texttt{v} has sequential consistency \\
\hline
\texttt{spawn [eternal] m $e$ [, $t$]} & \texttt{m} is a method,
$e$ is an expression, $t$ is the thread-local state \\
\hline
\texttt{trap m $e$} & \texttt{m} is a method and $e$ is an expression \\
\hline
\texttt{var $v$ = $e$} & $v$ is a new variable, $e$ is an expression\\
\hline
\texttt{when $b$: S} & $b$ is a boolean, \texttt{S} a list of statements\\
\hline
\texttt{when exists $a$ in $e$: S} & $a$ is a bound variable, $e$ is an expression\\
\hline
\texttt{while $b$: S} & $b$ is a boolean, \texttt{S} a list of statements\\
\hline
\end{tabular}

@@

\subsection*{\textbf{print}}

The statement \textbf{print $e$} atomically evaluates $e$ in read-only
mode and adds the result to the print log.  The print log is used to
create an ``external behavior DFA'' for the Harmony program.
